---
title: "Introduction to MultiOmicsCancer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to MultiOmicsCancer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction

**MultiOmicsCancer** is a modular, cancer-agnostic R package for integrative
multi-omics analysis in cancer research. It provides a unified framework for:

- Differential methylation analysis
- Differential gene expression analysis
- Methylation-expression integration
- Survival analysis
- Publication-quality visualization

The package works with **any cancer type** — from TCGA and cBioPortal to
user-provided datasets — without requiring code modifications.

# Installation

```{r install, eval = FALSE}
# Install from GitHub
devtools::install_github("maryam6770/MultiOmicsCancer")
```

```{r setup}
library(MultiOmicsCancer)
```

# Loading Data

The `load_cancer_data()` function loads multi-omics data from cBioPortal,
TCGA, or user-provided datasets. It returns a `MultiAssayExperiment` object
that all downstream functions can use.

## Option 1: Load from cBioPortal

```{r load-cbioportal}
# Load breast cancer data from cBioPortal
brca <- load_cancer_data(
  cancer_type = "brca_tcga",
  data_types = c("methylation", "expression")
)
```

## Option 2: Load custom data

If you have your own data, you can provide it as a named list of matrices.

```{r load-custom}
# Simulate custom multi-omics data
set.seed(42)
n_samples <- 20
n_genes <- 200

meth <- matrix(runif(n_genes * n_samples, 0, 1),
               nrow = n_genes, ncol = n_samples)
expr <- matrix(rnorm(n_genes * n_samples, mean = 8, sd = 1.5),
               nrow = n_genes, ncol = n_samples)
rownames(meth) <- paste0("Gene", seq_len(n_genes))
rownames(expr) <- paste0("Gene", seq_len(n_genes))
colnames(meth) <- paste0("Sample", seq_len(n_samples))
colnames(expr) <- paste0("Sample", seq_len(n_samples))

# Induce epigenetic silencing for the first 30 genes
for (i in 1:30) {
  expr[i, ] <- 15 - 5 * meth[i, ] + rnorm(n_samples, 0, 0.3)
}

custom_mae <- load_cancer_data(
  cancer_type = "my_study",
  data_types = c("methylation", "expression"),
  source = "custom",
  custom_data = list(methylation = meth, expression = expr)
)
```

# Differential Methylation Analysis

The `analyze_methylation()` function identifies hyper- and hypomethylated
probes between two groups.

```{r analyze-methylation, eval = FALSE}
dmr <- analyze_methylation(
  brca,
  group_column = "sample_type",
  group_levels = c("Solid Tissue Normal", "Primary Tumor"),
  delta_beta = 0.1,
  p_cutoff = 0.05
)

# View the top differentially methylated probes
head(dmr[dmr$significant, ])
```

# Differential Expression Analysis

The `analyze_expression()` function performs differential gene expression
analysis using either `limma` or `DESeq2`.

```{r analyze-expression, eval = FALSE}
deg <- analyze_expression(
  brca,
  method = "limma",
  logFC_cutoff = 1,
  p_cutoff = 0.05
)

# View the top differentially expressed genes
head(deg[deg$significant, ])
```

# Integrating Methylation and Expression

The `integrate_meth_expr()` function identifies genes whose expression is
significantly correlated with their promoter methylation.

```{r integrate}
integrated <- integrate_meth_expr(
  custom_mae,
  correlation_method = "pearson",
  cor_cutoff = 0.3,
  p_cutoff = 0.05
)

# Filter epigenetically silenced genes
silenced <- integrated[integrated$significant &
                        integrated$regulation == "Negative", ]
head(silenced)
```

# Survival Analysis

The `analyze_survival()` function performs Kaplan-Meier and Cox regression
analysis for any molecular feature.

```{r survival, eval = FALSE}
surv <- analyze_survival(
  brca,
  time_column = "OS.time",
  event_column = "OS",
  feature_name = "TP53",
  split_method = "median"
)

# View log-rank p-value
surv$logrank_p

# View Cox model summary
surv$cox_summary$coefficients
```

# Visualization

The `plot_cancer()` function creates publication-quality plots.

## Volcano plot of differential expression

```{r plot-volcano, eval = FALSE}
p1 <- plot_cancer(deg, plot_type = "volcano", fc_cutoff = 1)
print(p1)
```

## Correlation histogram

```{r plot-correlation}
p2 <- plot_cancer(integrated, plot_type = "correlation")
print(p2)
```

## Survival curve

```{r plot-survival, eval = FALSE}
p3 <- plot_cancer(plot_type = "survival", surv_result = surv)
print(p3)
```

# Session Information

```{r session}
sessionInfo()
```
